---
title: "BIOS6621 HW6"
author: "Ryan Summers"
date: "2024-10-01"
output:
  pdf_document: default
  html_document: default
---

```{r Libraries, message=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(magrittr)
library(kableExtra)
library(gt)
```


```{r Load Data File}
lead_df <- read_csv('/Users/ryan_summers/UC - M.S. Biostatistics/BIOS 6621 - Statistical Consulting/Week 6/HW 6/lead-iq-01.csv')

```

```{r}
lead_df %>%
  ggplot(aes(x = factor(Smelter), y = IQ)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~Smelter, scales = 'free') +
  theme_bw() +
  labs(title = "IQ Levels by Smelter", y = "IQ Levels")

```


```{r Table1, echo=FALSE}
# Summarizing the data
# sum_tbl <- lead_df %>%
#   group_by(Smelter) %>% 
#   summarise(
#     mean_iq = round(mean(IQ)),
#     q1_iq = quantile(IQ, probs=0.25),
#     median_iq = round(median(IQ)),
#     q3_iq = quantile(IQ, probs=0.75),
#     iqr_iq = IQR(IQ),
#     sd_iq = round(sd(IQ)),
#     min_iq = min(IQ),
#     max_iq = max(IQ) )
# 
# colnames(sum_tbl) <- c('Smelter', 'Mean', 'Q1', 'Median', 'Q3', 'IQR(Q3-Q1)', 'SD', 
#                        'Min', 'Max')
# 
# kable(sum_tbl, format = "html", 
#         caption = "<span style=
#         ' font-size: 20px; 
#           font-weight: bold; 
#           color: #333; '
#         >Summary Statistics</span>") %>%
#   add_header_above(c(" " = 1, "IQ" = 8)) %>% 
#   kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 


```


```{r Table 2}
lead_df %>%
  group_by(Smelter) %>% 
  summarise(
    n = n(),
    mean_iq = round(mean(IQ)),
    q1_iq = quantile(IQ, probs=0.25),
    median_iq = round(median(IQ)),
    q3_iq = quantile(IQ, probs=0.75),
    iqr_iq = IQR(IQ),
    sd_iq = round(sd(IQ)),
    min_iq = min(IQ),
    max_iq = max(IQ) ) %>% 
  
  # rowname_col = "row", groupname_col = "group"
  gt() %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  
  # modify row cells to be gray
  tab_style(
    style = cell_fill(color='gray95'),
    locations = list(
      cells_body(rows = 1),
      cells_title())) %>%

  # modify column names
  cols_label(
    n ~ 'n',
    mean_iq ~ "Mean",
    q1_iq ~ "Q1",
    median_iq ~ 'Median',
    q3_iq ~ 'Q3',
    iqr_iq ~ 'IQR',
    sd_iq ~ 'SD',
    min_iq ~ 'Min',
    max_iq ~ 'Max'
  ) %>% 
  
  # add a source note
  tab_source_note(
    source_note = md(c(
    '_Note:_ The original study was published in Landrigan PJ, Baloh RW, Barthel WF, 
    Whitworth RH, Staehling NW, and Rosenblum BF. (1975). Neuropsychological 
    dysfunction in children with chronic low-level lead absorption. The Lancet, 
    305: 708-712.',
    
    '_Data Source:_ Rosner B. (2006). Fundamentals of Biostatistics (6th edition). Duxbury.'))) %>%
  
  # add a footnote (if needed)
  # tab_footnote(
  #   footnote = "This is a footnote",
  #   locations = cells_column_labels(2)) %>%
  
  # add a header and subtitle
  tab_header(
    title = 'Table 1: Summary Statistics of IQ Levels',
    subtitle = "Children: Ages between 3 & 15 in El Paso") %>% 
  
  # adjust column widths of table
  tab_options(table.width = '85%') 

```

#### Analysis 

The summary statistics indicate that the IQ levels of children between the ages 
of 3 and 15 in El Paso were approximately normally distributed for children near 
the Smelter. However, an extreme outlier was noted for IQ levels in children far
from the Smelter. Given this outlier, the median and IQR values are better 
equipped to analyze the variability between groups. The medians were within five
points of each other, and the IQRs were approximately equally spread between the third (Q3) 
and first (Q1) quartiles, indicating very little difference between the groups.

#### Mean IQ

The mean IQ for children far from smelter:
`r round(mean(lead_df$IQ[lead_df$Smelter == 'Far']))`.

The mean IQ for children near the smelter: 
`r round(mean(lead_df$IQ[lead_df$Smelter == 'Near']))`.






